Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-28 resubmission/04-analisis.qmd
Up ajustes (renderizado)

Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average values across survey waves. The results show a notable upward trend in market justice preferences, particularly in the most recent waves. Perceived economic inequality exhibited the highest standardized mean in 2016; however, this variable shows a general downward trajectory over time, with a temporary increase in 2019, possibly reflecting the effects of the social uprising that year. Interestingly, while perceptions of inequality declined in the latest waves (2022–2023), market justice preferences continued to rise. Meritocratic beliefs, in contrast, remain relatively stable overall. Nevertheless, perceptions that individuals are rewarded based on talent tend to be slightly higher than those based on effort. Both meritocracy-related measures follow a similar temporal pattern: an increase from 2016 to 2018, a decline between 2019 and 2022—potentially associated with the social unrest and the consequences of the COVID-19 pandemic—and a subsequent rise in 2023, returning to levels comparable to those observed in 2016.

@@

When examining the between-group effects (BE) in Model 5 and 6, which capture differences between individuals in the average of the main variables, a similar pattern emerges. Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.002, $p$ > .05). However, this effects is no longer statistically significant. In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.206, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.036).

>>>>

Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average values across survey waves. The results show a notable upward trend in market justice preferences, particularly in the most recent waves. Perceived economic inequality exhibited the highest standardized mean in 2016; however, this variable shows a general downward trajectory over time, with a temporary increase in 2019, possibly reflecting the effects of the social uprising that year. Interestingly, while perceptions of inequality declined in the latest waves (2022–2023), market justice preferences continued to rise. Meritocratic perceptions, in contrast, remain relatively stable overall. Nevertheless, perceptions that individuals are rewarded based on talent tend to be slightly higher than those based on effort. Both meritocracy-related measures follow a similar temporal pattern: an increase from 2016 to 2018, a decline between 2019 and 2022—potentially associated with the social unrest and the consequences of the COVID-19 pandemic—and a subsequent rise in 2023, returning to levels comparable to those observed in 2016.

@@

When examining the between-group effects (BE) in Model 5 and 6, which capture differences between individuals in the average of the main variables, a similar pattern emerges. Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.002, $p$ > .05). However, this effect is no longer statistically significant. In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.206, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.036, $p$ > .05).

### 2025-07-28 resubmission/05-discussion.qmd
Up ajustes (renderizado)

This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic perceptions, and attitudes toward market justice in Chile. It posits that higher perceived inequality and stronger meritocratic perceptions both independently contribute to greater support for market-based distributions of social goods. Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic perceptions on market justice preferences. Additionally, it considers how major social and political mobilizations, such as those occurring around 2019 and subsequent years, may have weakened these relationships, reflecting increased critical perspectives on systemic inequalities and market mechanisms. The subsequent discussion interprets the empirical findings in light of these hypotheses, considering their implications for understanding attitudes toward inequality and social justice in a context marked by profound social change.

The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice. The findings demonstrated that individuals perceiving higher income disparities showed less support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021]. This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity. Similarly, the negative within-effects observed over time ($H1_{b}$) implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system. This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also further foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.

>>>>

This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile. It posits that higher perceived inequality and stronger meritocratic perceptions both independently contribute to greater support for market-based distributions of social goods. Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic perceptions on market justice preferences. Additionally, it considers how major social and political mobilizations, such as those occurring around 2019 and subsequent years, may have weakened these relationships, reflecting increased critical perspectives on systemic inequalities and market mechanisms. The subsequent discussion interprets the empirical findings in light of these hypotheses, considering their implications for understanding attitudes toward inequality and social justice in a context marked by profound social change.

The first hypothesis suggested that perceptions of economic inequality were strongly linked to preferences for market justice. The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021]. This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity. Similarly, the negative within-effects observed over time ($H1_{b}$) implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system.

### 2025-07-28 resubmission/06-conclusion.qmd
Up ajustes (renderizado)

This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and attitudes toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey. By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal, neoliberal environment.

The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health. At the same time, meritocratic beliefs—particularly those emphasizing individual effort—are strongly associated with support for market-based distributions, suggesting that these beliefs serve as moral justifications for structural inequalities. However, changes in meritocratic perceptions over time reveal a more nuanced picture: while increased emphasis on effort reinforces support for market justice, increased emphasis on talent tends to reduce it. This distinction likely reflects broader beliefs about the controllability and fairness of different meritocratic traits. Moreover, the interaction between inequality perceptions and meritocratic beliefs indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative—that is, when individuals are less aware or concerned about economic disparities. This finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.

This research advances the extant literature by integrating subjective perceptions with social and political contexts to explain attitudes toward economic inequality and distributional justice. While previous studies primarily focused on objective measures or individual characteristics, this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time. It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness. The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual perceptions and preferences.

Despite its contributions, the study has several limitations. First, the reliance on self-reported perceptions and attitudes may be subject to social desirability bias. Second, the focus on Chile, while offering valuable insights into a specific context, limits the generalizability of results to other countries with different institutional and cultural backgrounds. Third, the observational nature of the data prevents causal inferences, and unobserved confounding factors may influence the observed relationships. Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed. Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs. Comparative studies across different national contexts would help determine the universality or specificity of these dynamics. Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes. Finally, examining how these perceptions influence behavioral outcomes, such as political participation or support for social movements, would provide valuable insights into the pathways from beliefs to collective action and policy change.

>>>>

This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey. By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal and commoditized environment.

The findings confirm that higher perceptions of economic inequality are associated with less support for market justice preferences, this is, the belief that it is fair that those with higher income have better social services such as education, pensions and health. At the same time, meritocratic perceptions—particularly those emphasizing individual effort—are strongly associated with support for market-based distributions, suggesting that meritocracy serve as a moral justification for structural inequalities. However, changes in meritocratic perceptions over time reveal a more nuanced picture: while increased emphasis on effort reinforces support for market justice, increased emphasis on talent tends to reduce it. This distinction likely reflects broader beliefs about the controllability and fairness of different meritocratic traits. Moreover, the interaction between inequality perceptions and meritocratic perceptions indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative—that is, when individuals are less aware or concerned about economic disparities. This finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.

This research advances the extant literature by integrating subjective perceptions with social and political contexts to explain attitudes toward economic inequality and distributional justice. While previous studies primarily focused on objective measures or individual characteristics [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015], this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time. It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness. The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual perceptions and preferences.

Despite its contributions, the study has several limitations. First, the reliance on self-reported perceptions and attitudes may be subject to social desirability bias. Second, the focus on Chile, while offering valuable insights into a specific context, limits the generalizability of results to other countries with different institutional and cultural backgrounds. Third, the observational nature of the data prevents causal inferences, and unobserved confounding factors may influence the observed relationships. Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed.

Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs. Comparative studies across different national contexts would help determine the universality or specificity of these dynamics. Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes. Finally, examining how these perceptions influence behavioral outcomes, such as political participation or support for social movements, would provide valuable insights into the pathways from beliefs to collective action and policy change.
